Mapping and matrix decomposition for biomedical images

### 2016-07-13 antsr/antsrdemo.Rmd
ENH: quick demo of some features

@@

![images at measurement fields](figures/statisticalfields.png)

@@

## Agnostic statistics
![](./figures/antsrex.jpg)

@@

This is an executable *ANTsR* code block - *N*-dimensional statistics to go with
our *N*-dimensional image processing software!
```{r loadimages,message=FALSE,warning=FALSE,results='hide'}
library(ANTsR)
dim<-2
filename<-getANTsRData('r16')
img<-antsImageRead( filename , dim )
filename<-getANTsRData('r64')
img2<-antsImageRead( filename , dim )
mask<-getMask(img,50,max(img),T)
mask2<-getMask(img,150,max(img),T)
nvox<-sum( mask == 1 )
nvox2<-sum( mask2 == 1 )
```
The brain has `r nvox` voxels ...

@@

Simulate a population morphometry study - a "VBM" ...
```{r morph,results='hide'}
simnum<-10
imglist<-list()
imglist2<-list()
for ( i in 1:simnum ) {
  img1sim<-antsImageClone(img)
  img1sim[ mask==1 ]<-rnorm(nvox,mean=0.5)
  img1sim[ mask2==1 ]<-rnorm(nvox2,mean=2.0)
  img2sim<-antsImageClone(img2)
  img2sim[ mask==1 ]<-rnorm(nvox,mean=0.20)
  imglist<-lappend(imglist,img1sim)
  imglist2<-lappend(imglist2,img2sim)
}
imglist<-lappend( imglist, imglist2 )
mat<-imageListToMatrix( imglist, mask )
DX<-factor( c( rep(0,simnum), rep(1,simnum) ) )
mylmresults<-bigLMStats( lm( mat ~ DX ) )
qvals<-p.adjust( mylmresults$pval.model )
```
The minimum q-value is `r min(qvals)` ...

@@

## Visualize the anatomical distribution
```{r vizmorph2hide,results='hide',echo=FALSE}
betas<-antsImageClone(mask)
betavec<-abs(mylmresults$beta.t[1,])
betas[mask==1]<-betavec
smoothImage(betas, 1.5)
thresh<-paste(3.0,'x',max(betavec),sep='')
ofn<-paste(figpath,'vizmorph2.jpg',sep='')
```

```{r vizmorph2,results='hide'}
plot.antsImage(img,functional=list(betas),threshold=thresh,
  outname=ofn)
```

![](`r ofn`)

## Network visualization
see `?plotBasicNetwork`
```{r networkfn,results='hide',warning=FALSE,echo=FALSE,message=FALSE}
ofn<-paste(figpath,'network',sep='')
ofnpng<-paste(ofn,'.png',sep='')
```
```{r network,results='hide',warning=FALSE,echo=FALSE,message=FALSE,eval=F}
       mnit<-getANTsRData("mni")
       mnit<-antsImageRead(mnit,3)
       mnia<-getANTsRData("mnia")
       mnia<-antsImageRead(mnia,3)
       ThresholdImage(3,mnit,mnit,1,max(mnit))
       ImageMath(3,mnit,"FillHoles",mnit)
       cnt<-getCentroids( mnia, clustparam = 50 )
       aalcnt<-cnt$centroids[1:90,]
       brain<-renderSurfaceFunction( surfimg =list( mnit ) , alphasurf=0.1 ,smoothsval = 1.5 )
       testweights<-matrix( rep( 0, 90*90 ) ,nrow=90)
       testweights[31,37]<-1  # ant cingulate to hipp
       testweights[31,36]<-2  # ant cingulate to post cingulate
       testweights[11,65]<-3  # broca to angular
       plotBasicNetwork( centroids = aalcnt , brain , weights=testweights )
       id<-par3d("userMatrix")
       rid<-rotate3d( id , -pi/2, 1, 0, 0 )
       rid2<-rotate3d( id , pi/2, 0, 0, 1 )
       rid3<-rotate3d( id , -pi/2, 0, 0, 1 )
       par3d(userMatrix = id )
       dd<-make3ViewPNG(  rid, id, rid2,  ofn )
       par3d(userMatrix = id )
```
![](`r ofnpng`)

@@

![](./figures/antsgoals.png)

>>>>

---
title: "Mapping and matrix decomposition for biomedical images"
author: "Brian B. Avants"
date: "July 11, 2016"
output: ioslides_presentation
---

@@

```{r, out.width = 800, fig.retina = NULL, echo=FALSE}
knitr::include_graphics("~/data/antsExamples/ANTs2015/figures/statisticalfields.png")
```

@@

## Agnostic statistics

```{r, out.width = 800, fig.retina = NULL, echo=FALSE}
knitr::include_graphics("/Users/bavants/data/antsExamples/ANTs2015/figures/antsrex.jpg")
```

@@

This is an executable *ANTsR* code block - *N*-dimensional statistics to go with
our *N*-dimensional image processing software!
```{r loadimages,message=FALSE,warning=FALSE,results='hide'}
library( ANTsR )
library( ggplot2 )
filename<-getANTsRData('r16')
img<-antsImageRead( filename )
filename2<-getANTsRData('r16')
img2<-antsImageRead( filename2 )
mask<-getMask( img,  50, max(img), cleanup = 5 )
mask2<-getMask( img, 150, max(img), cleanup = 5 )
nvox<-sum( mask == 1 )
nvox2<-sum( mask2 == 1 )
img3<-antsImageRead( getANTsRData('r27') )
mask3<-getMask( img3,  200, max(img3), cleanup = 5 )
nvox3 = sum( mask3 == 1 )
```
The brain has `r nvox` voxels ...

## Run a diffeomorphic mapping

```{r qreg,message=FALSE,warning=FALSE,results='hide'}
plot( img )
img64 = antsImageRead( getANTsRData( 'r64' ) )
plot( img64 )
areg = antsRegistration( img, img64, typeofTransform = 'Affine' )
plot( areg$warpedmov )
sreg = antsRegistration( img, img64, typeofTransform = 'SyN' )
plot( sreg$warpedmov )
jacobian = createJacobianDeterminantImage( img, sreg$fwdtransforms[[1]], 1 )
plot( jacobian )
plot( img, jacobian * -1 , window.overlay = c(0.2,1) )
plot( img, sreg$warpedmovout  %>% iMath("Canny", 5, 0.1, 1) )
```

## Run a population diffeomorphic mapping

```{r qregpop,message=FALSE,warning=FALSE,results='hide'}
ilist = list( 
  antsImageRead( getANTsRData( 'r27') ),
  antsImageRead( getANTsRData( 'r30') ),
  antsImageRead( getANTsRData( 'r62') ),
  antsImageRead( getANTsRData( 'r64') ),
  antsImageRead( getANTsRData( 'r85') ) )
rlist = list( )
for ( i in 1:length( ilist ) ) {
  temp = antsRegistration( img, ilist[[i]], typeofTransform = "SyN" )
  jacobian = createJacobianDeterminantImage( img, temp$fwdtransforms[[1]], 1)
  rlist[[ i ]] = jacobian }
imask = getMask( img )
mat   = imageListToMatrix( rlist, imask )
eanat = eanatDef( scale( mat ), nvecs = esel, mask = mask,
  positivity = TRUE, smoother = 0.5, cthresh = 200 )
```

## Run a population eigenanatomy

```{r qreanat,message=FALSE,warning=FALSE,results='hide'}
eanat = eanatDef( scale( mat ), nvecs = 4, mask = imask,
  positivity = TRUE, smoother = 0.5, cthresh = 2200 )
eilist = matrixToImages( eanat, imask )
eseg = eigSeg( imask, eilist )
plot( img, eseg )
```

@@

Simulate a population morphometry study - a "VBM" ...
```{r morph,results='hide'}
simnum<-10
imglist<-list()
imglist2<-list()
for ( i in 1:simnum ) {
  img1sim<-antsImageClone(img)
  img1sim[ mask==1 ]  = rnorm(  nvox, mean=1.0 )
  img1sim[ mask2==1 ] = rnorm( nvox2, mean=1.5 )
  img2sim = antsImageClone(img2)
  img2sim[ mask==1 ] = rnorm( nvox, mean=1.0)
  img2sim[ mask3==1 ] = rnorm( nvox3, mean=0.0)
  imglist<-lappend(  imglist ,  img1sim %>% smoothImage( 5 ) )
  imglist2<-lappend( imglist2,  img2sim %>% smoothImage( 5 ) )
}
imglist<-lappend( imglist, imglist2 )
```

Simulate a population morphometry study - a "VBM" ...
```{r morph2,results='hide'}
mat<-imageListToMatrix( imglist, mask )
DX<-factor( c( rep(1,simnum), rep(0,simnum) ) )
mylmresults<-bigLMStats( lm( mat ~ DX ) )
qvals<-p.adjust( mylmresults$pval.model )
```
The minimum q-value is `r min(qvals)` ...

@@

## Visualize the anatomical distribution
```{r vizmorph2hide,results='hide',echo=FALSE}
betas<-antsImageClone(mask)
betavec<-( mylmresults$beta.t[1,] )
betas[mask==1]<-betavec
thresh<-c( 5.0, max(betavec) )
plot( img, betas, window.overlay=thresh )
```

## Eigenanatomy: Run it

```{r eanat}
esel = eanatSelect( scale( mat ), mask, 200, 0.5, maxNEvec = 10 )
eanat = eanatDef( scale( mat ), nvecs = esel, mask = mask,
  positivity = TRUE, smoother = 0.5, cthresh = 200 )
```

## Eigenanatomy: Display it

```{r eanatd}
mm = makeImage( mask, eanat[1,] ) %>% iMath("Normalize")
plot( img, mm, window.overlay=c(0.1,1) )
```

## Test the eigenanatomy components

The first component should relate to a "group difference."

Let's just use linear regression.

```{r eanatregression}
dx = factor( c( rep(1, (nrow(mat)/2) ), rep(0, (nrow(mat)/2) )   ) )
# make the low-dimensional representation
proj = mat %*% t( eanat )
eanatDataFrame = data.frame( dx = dx, eanat = proj )
mdl = lm( eanat.1 ~ dx , data = eanatDataFrame )
print( knitr::kable( coefficients( summary( mdl ) ) ) )
mdl = lm( eanat.5 ~ dx , data = eanatDataFrame )
print( knitr::kable( coefficients( summary( mdl ) ) ) )
```

## A "real study"

## Network visualization
see `?plotBasicNetwork`
```{r networkfn,results='hide',warning=FALSE,echo=FALSE,message=FALSE}
ofn<-paste('/Users/bavants/data/antsExamples/ANTs2015/figures/network',sep='')
ofnpng<-paste(ofn,'.png',sep='')
```

```{r network,results='hide',warning=FALSE,echo=FALSE,message=FALSE,eval=F}
       mnit<-getANTsRData("mni")
       mnit<-antsImageRead(mnit,3)
       mnia<-getANTsRData("mnia")
       mnia<-antsImageRead(mnia,3)
       ThresholdImage(3,mnit,mnit,1,max(mnit))
       ImageMath(3,mnit,"FillHoles",mnit)
       cnt<-getCentroids( mnia, clustparam = 50 )
       aalcnt<-cnt$centroids[1:90,]
       brain<-renderSurfaceFunction( surfimg =list( mnit ) , alphasurf=0.1 ,smoothsval = 1.5 )
       testweights<-matrix( rep( 0, 90*90 ) ,nrow=90)
       testweights[31,37]<-1  # ant cingulate to hipp
       testweights[31,36]<-2  # ant cingulate to post cingulate
       testweights[11,65]<-3  # broca to angular
       plotBasicNetwork( centroids = aalcnt , brain , weights=testweights )
       id<-par3d("userMatrix")
       rid<-rotate3d( id , -pi/2, 1, 0, 0 )
       rid2<-rotate3d( id , pi/2, 0, 0, 1 )
       rid3<-rotate3d( id , -pi/2, 0, 0, 1 )
       par3d(userMatrix = id )
       dd<-make3ViewPNG(  rid, id, rid2,  ofn )
       par3d(userMatrix = id )
```

```{r, out.width = 800, fig.retina = NULL, echo=FALSE}
knitr::include_graphics( ofnpng )
```

@@

```{r, out.width = 800, fig.retina = NULL, echo=FALSE}
knitr::include_graphics("~/data/antsExamples/ANTs2015/figures/antsgoals.png")
```
